Approximation as Differentiation: Notes

### 2025-07-07 main/approx-as-tangents.tex
Remove duplicate definition of Galois connection.
[ tex ] flesh out the approximations as tangents section

The general study of differentiable functions takes place on \emph{manifolds}, which are topological spaces that ``locally'' look like a subset of the Euclidean space $\RR^n$. The spaces $\RR^n$ themselves are manifolds, but so are ``non-flat'' examples such as $n$-spheres and yet more exotic spaces. Every point $x$ in a manifold $M$ has an associated \emph{tangent vector space} $\tangents_x(M)$ and \emph{cotangent vector space} $\cotangents_x(M)$, the latter consisting of linear functions $\tangents_x(M) \linearto \RR$. The tangent and cotangent spaces are finite dimensional, so in the presence of a chosen basis they are canonically isomorphic. In the case when the manifold is $\RR^n$, then every tangent space is isomorphic to $\RR^n$ as well.

Smooth functions $f$ between manifolds $M$ and $N$ are functions on their points that are locally differentiable on $\RR^n$. Manifolds and smooth functions form a category $\Man$. Each smooth function induces maps of the (co)tangent spaces:
\begin{itemize}
\item The \emph{forward derivative} (tangent map or pushforward) $\pushf{f}_x$ is a linear map $\tangents_x(M) \linearto \tangents_{f(x)}(N)$. In the Euclidean case when $M = \RR^m$ and $N = \RR^n$, the tangent map can be represented by the Jacobian matrix of partial derivatives of $f$ at $x$.
\item The \emph{backward derivative} (cotangent map, or pullback) $\pullf{f}_x$ is a linear map $\cotangents_{f(x)}(N) \linearto \cotangents_x(M)$. In the Euclidean case, the backward derivative is represented by the transpose of the Jacobian of $f$ at $x$.
\end{itemize}

@@

\subsection{Stable Domain Theory as Differentiability}

Our thesis is that \GPS is a form of differentiable programming, where tangents are not linear approximations of functions but instead are qualitative approximations of elements in the sense of domain theory. Smooth functions in this setting are Berry's \emph{stable functions} \cite{berry79,berry82}. We now introduce these concepts and how they relate to \GPS.

@@

Domain theory is a method for defining the semantics of programs that handle infinite data (e.g., functions or infinite streams). \emph{Domains} are certain partially ordered sets where the ordering denotes a relationship of qualitative information content: if $x \sqsubseteq y$, then $y$ may contain more information than $x$. For example, if $x$ and $y$ are functions, then $y$ may be defined for more values in its domain than $x$. Infinite objects are understood in terms of their approximations in this sense, and domains are assumed to be closed under least upper bounds (lubs) of directed sets, meaning that any internally consistent collection of elements has a ``completion'' that contains all the information covered by the set. Programs are interpreted as monotone functions that preserve directed lubs. Monotonicity captures the idea that if the input gets more defined, then the output can get more defined. Preservation of lubs, or \emph{continuity}, states that a function interpreting a program cannot act inconsistently on approximations and their completion, which corresponds to the intuitive idea that a function that is computable cannot look at a non-finite amount of input to produce a finite output.

For the purposes of \GPS, we are not interested in using approximations to model computation on infinite objects, but instead to use them for the related use of revealing how programs explore their inputs when producing parts of their output. Therefore, we ignore completeness properties of the partially ordered sets we consider.

@@

When giving a denotational semantics for sequential programming languages, such as PCF, continuous functions are too permissive. Plotkin's Parallel OR (\exref{parallel-or}, below) is continuous but does not explore its input in a way consistent with a sequential implementation \cite{plotkin77lcf}. Continuous functions can explore their input in a ``random access'' way, which is incompatible with the central idea in \GPS that we should be able to identify a \emph{minimal} part of the input that leads to a part of the output.

Stability is a property that can be required of monotone functions that was invented by \citet{berry79} in an attempt to capture sequentiality. This was not successful (see the $\mathrm{gustave}$ function in \exref{parallel-or}), but we will see now how it is closely related to problem of computing forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet{amadio-curien} (Chapter 12). We start with a property that is weaker than stability, but is easier to motivate in connection with derivatives of smooth functions.

In a partially ordered set $X$, for any $x \in X$ the set of elements below $x$, $\downset{x} = \{ x' \mid x' \sqsubseteq x \}$, is itself a partially ordered set. These approximations of $x$ we will think of as ``derivatives'' of $x$, and the whole set $\downset{x}$ as the ``tangent space''. Tangent spaces are vector spaces, since they are linear approximations to curves at that point. In the partially ordered setting, we take elements of $\downset{x}$ to be approximations of processes defined at $x$. As we can add tangents, we assume we can take meets of approximations in $\downset{x}$:

@@

Smooth functions have a ``pushforward'' derivative that takes tangents at a point $x$ to tangents at the point $f(x)$ and is linear. In the partially ordered setting, we are going to take a function $f$'s pushforward derivative at $x$ to be its restriction to $\downset{x}$, taking approximations of the input to approximations of the output. Matching the linearity of the pushforward derivative smooth functions, we require that these restrictions preserve meets:

@@

\begin{lemma}
  Morphisms of meet approximation spaces are closed under identities
  and composition, forming a category. FIXME: with products?
  functions? sums?
\end{lemma}

\begin{example}[Conditionally Multiplicative Functions]
\label{ex:strict-short-circuit}
  To see the effect of requiring preservation of bounded meets,
  consider several ways of defining disjunction on the lifted booleans
  $\mathbb{B}_\bot$. Two functions that are cm are the strict and
  left-short-circuiting ORs\footnote{The clauses in these examples are
    shorthand for the graph of the function. They are not to be
    understood as pattern matching clauses in a language like Haskell,
    where it is not possible to match on $\bot$.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
    \qquad
    \raisebox{-15ex}{
      \begin{tikzpicture}
        \node (top) at (0,0) {$(\mathsf{tt},\mathsf{ff})$};
        \node [below of=top,xshift=-0.5cm] (oi) {$(\bot,\mathsf{ff})$};
        \node [below of=top,xshift=0.5cm] (io) {$(\mathsf{tt},\bot)$};
        \node [below of=oi,xshift=0.5cm] (oo) {$(\bot,\bot)$};
        \draw (top) -- (oi);
        \draw (top) -- (io);
        \draw (oi) -- (oo);
        \draw (io) -- (oo);
      \end{tikzpicture}
    }
  \end{displaymath}
  In the poset $\mathbb{B}_\bot^2$, a typical poset of approximations of a fully defined element is shown to the right. For $\mathrm{strictOr}$, any approximation that isn't the fully defined input is mapped to $\bot$, while $\mathrm{shortCircuitOr}$ maps the partially defined $(\mathsf{tt},\bot)$ to $\mathsf{tt}$. Thus, even though these functions operate identically on fully defined inputs, they differ in their \emph{derivatives} on partially defined input, exposing how . That they are cm can be checked by examining their restrictions' behaviour. If we take the approximations $(\bot,\mathsf{ff})$ and $(\mathsf{tt},\bot)$, then their meet is $(\bot,\bot)$; $\mathrm{strictOr}$ maps all three elements to $\bot$, so is cm here since $\bot \wedge \bot = \bot$; and $\mathrm{shortCircuitOR}$ has $(\bot,\mathsf{ff}) \mapsto \bot$ and $(\mathsf{tt},\bot) \mapsto \mathsf{tt}$, the meet of which is $\bot = \mathrm{shortCircuitOR}(\bot,\bot)$. Other combinations can be checked similarly.
\end{example}

@@

Conditional multiplicativity seems to be a reasonable analogue to functions with a well-defined notion of derivative. However, for \GPS we also require an analogue to the reverse derivative, where we map approximations backwards to give the least approximation of the input for a given approximation of the output. In the case of smooth functions, we are always guaranteed a reverse derivative but there is not always a best way to map approximations backwards for cm functions, as the following example shows.

\begin{example}[Is Conditional Multiplicativity Enough?]
  \label{ex:non-stable-function}
  % In the preceeding examples, we demonstrated conditional
  % multiplicativity by giving a backwards mapping of approximations
  % that forces preservation of meets. A natural question is whether or
  % not we always have such a backwards map which we could use for
  % backwards propagation of approximations.
  An example that is conditionally multiplicative, but does not admit a backards map of approximations (from \citet{amadio-curien} just before Lemma 12.2.3, originally due to Berry) is given by defining $\mathrm{unstable} : D \to \{\bot, \top\}$, where:
  \begin{displaymath}
    D = \bot \sqsubseteq \cdots \sqsubseteq n \sqsubseteq \cdots \sqsubseteq 1 \sqsubseteq 0
  \end{displaymath}
  as $\mathrm{unstable}(\bot) = \bot$ and $\mathrm{unstable}(n) = \top$. This is monotone, and preserves meets in every $\downset{x}$. However, there is no ``best'' (i.e., least) input that gives us any finite output.
\end{example}

@@

In light of the preceeding example, we can give an alternative
definition of a function that requires the existence of a reverse
mapping directly, even without assuming that any meets exist:

@@

  \item As discussed above, $\mathrm{unstable}$ is not stable, but is conditionally multiplicative.
  \end{enumerate}
\end{example}

The fact that these two functions' stability witnesses reveal information about how they depend on their input is what we will exploit in order to use the idea of stability for \GPS. It is interesting that sequentiality is not necessarily required in order to obtain backwards maps of approximations, as might be assumed by thinking about how to trace computations for provenance information.

Stability has an alternative definition in terms of Galois connections, which will be more useful for what follows. This characterisation is due to~\citet{taylor99}. We first formally define Galois connections, which we used informally in \exref{introduction-example}.

\begin{definition}[Galois connection]
  \label{def:galois-connection}
  Suppose $X$ and $Y$ are posets. A \emph{Galois connection} $f \adj g: X \to Y$ is a pair of monotone functions $f: Y \to X$ and $g: X \to Y$ satisfying $y \leq g(x) \iff f(y) \leq x$ for any $x \in X$ and $y \in Y$. We refer to $f$ as the left adjoint and $g$ as the right adjoint.
\end{definition}

@@

Even though stable functions can be defined on any partially ordered set, in light of the analogy with tangent spaces, it makes sense to require that meets, preserved by forward approximation maps, and joins, preserved by backwards approximation maps, exist:

@@

We could have required that L-posets' principal downsets were complete lattices, which would mean that we would only need to require that forward approximation maps preserve all meets and we could compute the backwards map automatically. However, our goal is a computable theory for \GPS, analogous to automatic differentiation. Requiring completeness to compute backwards maps of approximations would require that complete joins are always computable, which will not be the case for approximations of functions.

\begin{proposition}
  L-posets and stable functions form a category. \bob{and the main point is that composition of stability witnesses looks like a chain rule!}
\end{proposition}

\bob{L-poset is has products and coproducts, but is not closed. To be closed would require each $\downset{x}$ to be a \emph{complete} lattice. However, this is tricky to implement as we need an infinitary operation to compute the reverse approximation map for function types. See \cite{amadio-curien} for the proof that bounded-complete lattices with stable functions form a cartesian closed category.}

@@

Let us summarise our analogy between differentiable functions between manifolds and stable functions between L-posets:

\begin{itemize}
\item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}

\bob{Put a conjecture about Tangent Categories in somewhere}

Given this close connection between \GPS and differentiable programming, we can take structures intended for modelling automatic differentiation and use them to model \GPS. This will enable us to generalise and expand the scope of \GPS to act as a foundation for data provenance in a wide range of computational settings.

>>>>

The general study of differentiable functions takes place on
\emph{manifolds}, topological spaces that ``locally'' behave like an
open subset of the Euclidean space $\RR^n$. The spaces $\RR^n$
themselves are manifolds, but so are ``non-flat'' examples such as
$n$-spheres and yet more exotic spaces. Every point $x$ in a manifold
$M$ has an associated \emph{tangent vector space} $\tangents_x(M)$
consisting of linear approximations of curves on the manifold passing
through $x$. Each point also has a \emph{cotangent vector space}
$\cotangents_x(M) = \tangents_x(M) \linearto \RR$. The tangent and
cotangent spaces are finite dimensional, so in the presence of a
chosen basis they are canonically isomorphic. In the case when the
manifold is $\RR^n$, then every tangent space is isomorphic to $\RR^n$
as well.

Smooth functions $f$ between manifolds $M$ and $N$ are functions on their points that are locally differentiable on $\RR^n$. Manifolds and smooth functions form a category $\Man$. Each smooth function induces maps of the (co)tangent spaces:
\begin{itemize}
\item The \emph{forward derivative} (tangent map, pushforward) $\pushf{f}_x$ is a linear map $\tangents_x(M) \linearto \tangents_{f(x)}(N)$. In the Euclidean case when $M = \RR^m$ and $N = \RR^n$, the tangent map can be represented by the Jacobian matrix of partial derivatives of $f$ at $x$.
\item The \emph{backward derivative} (cotangent map, pullback) $\pullf{f}_x$ is a linear map $\cotangents_{f(x)}(N) \linearto \cotangents_x(M)$. In the Euclidean case, the backward derivative is represented by the transpose of the Jacobian of $f$ at $x$.
\end{itemize}

\bob{mention chain rule here? and tangent spaces?}

@@

\subsection{Stable Functions as Differentiable Functions}

Our thesis is that \GPS is a generalised form of differentiable programming, where tangents are not linear approximations of curves but instead are qualitative information approximations of elements. Smooth functions in this setting are Berry's \emph{stable functions} \cite{berry79,berry82}. We now introduce these concepts and how they relate to \GPS.

@@

Domain theory is a method for defining the semantics of programs that
handle infinite data such as functions or infinite streams. Domains
are certain partially ordered sets where the ordering denotes a
relationship of qualitative information content: if $x \sqsubseteq y$,
then $y$ may contain more information than $x$. For example, if $x$
and $y$ are functions, then $y$ may be defined at more values than
$x$. Infinite objects are understood in terms of their approximations
in this sense, and domains are assumed to be closed under least upper
bounds (lubs) of directed sets, meaning that any internally consistent
collection of elements has a ``completion'' that contains all the
information covered by the set. Programs are interpreted as monotone
functions that preserve directed lubs. Monotonicity captures the idea
that if the input gets more defined, then the output can get more
defined. Preservation of lubs, or \emph{continuity}, states that a
function interpreting a program cannot act inconsistently on
approximations and their completion, which corresponds to the
intuitive idea that a function that is computable cannot look at a
non-finite amount of input to produce a finite output. \cite{abramsky-jung} is a comprehensive introduction to domain theory.

For the purposes of \GPS, we are not interested in using
approximations to model computation on infinite objects, but instead
for revealing how programs explore their inputs when producing parts
of their output. Therefore, we ignore completeness properties of the
partially ordered sets we consider\footnote{\bob{we might want to bring this back when considering general recursive programs. Are bidomains \cite{berry79} useful? they have two orderings. See also Laird's Bistable Biorders.}}.

@@

When giving a denotational semantics for sequential programming
languages Scott-continuous functions are too permissive. Famously,
\citet{plotkin77lcf}'s Parallel OR (\exref{parallel-or}, below) is
continuous but does not explore its input in a way consistent with a
sequential implementation. Continuous functions can explore their
input in a non-deterministic way as long as the result is
deterministic. This non-determinism results in functions that whose
output cannot be assigned a unique minimal input that caused
it. Therefore, continuous functions are incompatible with the central
idea in \GPS that we should be able to identify a \emph{minimal} part
of the input that leads to a part of the output.

\emph{Stability} is a property that can be required of monotone functions that was invented by \citet{berry79} in an attempt to capture sequentiality. This was not successful (see the $\mathrm{gustave}$ function in \exref{parallel-or}), but we will see now how it is closely related to problem of computing forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet{amadio-curien} (Chapter 12). We start with a property that is weaker than stability, but is easier to motivate in connection with derivatives of smooth functions.

In a partially ordered set $X$, for any $x \in X$ the set of elements below $x$, $\downset{x} = \{ x' \mid x' \sqsubseteq x \}$, is itself a partially ordered set. These approximations of $x$ we will think of as ``tangents'' at $x$, and the whole set $\downset{x}$ as the ``tangent space''\footnote{\bob{Is it possible to derive the tangent spaces by thinking about continuous functions passing through $x$?}}. Tangent spaces are vector spaces, so in the partially ordered setting we take elements of $\downset{x}$ to be approximations of processes defined at $x$. As we can add tangents, we assume we can take meets of approximations in $\downset{x}$:

@@

For the approximation version of the forward derivative of $f$ at $x$, we take $f$'s restriction to $\downset{x}$, taking approximations of the input to approximations of the output. Matching the linearity of the forward derivative of smooth functions, we require that these restrictions preserve meets. This is exactly the definition of \emph{conditionally multiplicative} function from \citet{berry79}:

@@

\begin{theorem}
  Bounded meet lattices and conditionally multiplicatve functions form
  a category $\CM$. This category has products, coproducts and
  exponentials.
\end{theorem}

\begin{proof}
  \AGDA. See \citet{amadio-curien} for the case when the posets are
  also cpos. The crucial technical step, identified by
  \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq g$ iff $\forall x. f(x) \subseteq g(x)$) but instead
  the stable ordering $f \sqsubseteq g$ iff \bob{FIXME}.
\end{proof}

\begin{example}[Conditionally Multiplicative Functions]
  \label{ex:strict-short-circuit}
  To see the effect of conditional multiplicativity, consider several
  ways of defining the OR on the lifted booleans
  $\mathbb{B}_\bot$. Two functions that are cm are the strict and
  left-short-circuiting ORs\footnote{The clauses in these examples are
    shorthand for the graph of the function. They are not to be
    understood as pattern matching clauses in a language like Haskell,
    where it is not possible to match on $\bot$.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
    \qquad
    \raisebox{-15ex}{
      \begin{tikzpicture}
        \node (top) at (0,0) {$(\mathsf{tt},\mathsf{ff})$};
        \node [below of=top,xshift=-0.5cm] (oi) {$(\bot,\mathsf{ff})$};
        \node [below of=top,xshift=0.5cm] (io) {$(\mathsf{tt},\bot)$};
        \node [below of=oi,xshift=0.5cm] (oo) {$(\bot,\bot)$};
        \draw (top) -- (oi);
        \draw (top) -- (io);
        \draw (oi) -- (oo);
        \draw (io) -- (oo);
      \end{tikzpicture}
    }
  \end{displaymath}
  In the poset $\mathbb{B}_\bot^2$, a typical poset of approximations of a fully defined element is shown to the right. For $\mathrm{strictOr}$, any approximation that isn't the fully defined input is mapped to $\bot$, while $\mathrm{shortCircuitOr}$ maps the partially defined $(\mathsf{tt},\bot)$ to $\mathsf{tt}$. Thus, even though these functions operate identically on fully defined inputs, they differ in their \emph{derivatives} on partially defined input, exposing how . That they are cm can be checked by examining their restrictions' behaviour. If we take the approximations $(\bot,\mathsf{ff})$ and $(\mathsf{tt},\bot)$, then their meet is $(\bot,\bot)$; $\mathrm{strictOr}$ maps all three elements to $\bot$, so is cm here since $\bot \wedge \bot = \bot$; and $\mathrm{shortCircuitOR}$ has $(\bot,\mathsf{ff}) \mapsto \bot$ and $(\mathsf{tt},\bot) \mapsto \mathsf{tt}$, the meet of which is $\bot = \mathrm{shortCircuitOR}(\bot,\bot)$. Other combinations can be checked similarly.
\end{example}

@@

\begin{example}[Intervals and Maximal Elements]
  \label{ex:intervals-and-maxima-elements}
  The set of intervals
  $\{ [l,u] \in \mathbb{R} \times \mathbb{R} \mid l \leq u \}$ ordered
  by reverse inclusion forms a (Scott) domain
  \cite{scott-outline}. The set of maximal elements is exactly
  $\mathbb{R}$. This domain has been proposed as a model of
  approximate real number computation \cite{escardo-real-pcf}. The
  information approximation reading is intuitive: as intervals move up
  the order they become tighter, containing more information about the
  number they are approximating.

  Given this reading, it makes sense to wonder if we can use interval
  approximations as ``information tangents'' of real numbers, where
  derivatives take approximating intervals to approximating
  intervals. Since intervals form a Scott domain, they are bounded
  complete and hence have bounded meets. However, the addition
  function on intervals, $[l_1,u_1] + [l_2,u_2] = [l_1+l_2,u_1+u_2]$,
  is not conditionally multiplicative, as can be easily checked.

  A solution is to use the set of intervals with nominated points that
  they are approximating: $\{[l,x,u] \mid l \leq x \leq u\}$. The
  ordering now is that $[l_1,x_1,u_1] \sqsubseteq [l_2,x_2,u_2]$ iff
  $x_1 = x_2$ and $l_1 \leq l_2$ and $u_2 \leq u_1$. Consequently, the
  maximal elements are $[x,x,x]$, recovering $\mathbb{R}$ again, but
  the approximations of each number all form independent sub-lattices.
  Addition is defined as
  $[l_1,x_1,u_1] + [l_2,x_2,u_2] = [(l_1+x_2) \sqcap
  (l_2+x_1),x_1+x_2,(u_1+x_2)\sqcup(u_2+x_1)]$, which is conditionally
  multiplicative. Note how this definition bears a resemblance to the
  product rule for derivatives, with (in the lower end of the
  interval) $\sqcap$ replacing $+$.

  This example is important to us because it shows that (for total
  programs) we are separately interested in the maximal elements and
  their approximations, and that approximations of each maximal
  element may have to be considered separately.

  Relatedly, \citet{edalat-heckmann98} proposed a Scott domain of
  formal balls on a metric space, where again the maximal elements are
  the points of the original space. More generally,
  \citet{continuous-lattice-book} (Section V-6) describes \emph{Domain
    environments}, which are domains whose maximal elements are
  exactly the points of a topological space. We are not aware of any
  work linking domain environments in general to stable functions. It
  would be interesting to see whether the situation for intervals,
  where approximations must be relative to a nominated point, repeats
  in general when considering conditionally multiplicative functions.
\end{example}

Given these examples, conditional multiplicativity seems to be a reasonable analogue to functions with a well-defined notion of derivative. However, for \GPS we also require an analogue to the reverse derivative, where we map approximations backwards to give the least approximation of the input for a given approximation of the output. In the case of smooth functions, we are always guaranteed a reverse derivative but there is not always a best way to map approximations backwards for cm functions, as the following example shows.

\begin{example}[Is Conditional Multiplicativity Enough?]
  \label{ex:non-stable-function}
  % In the preceeding examples, we demonstrated conditional
  % multiplicativity by giving a backwards mapping of approximations
  % that forces preservation of meets. A natural question is whether or
  % not we always have such a backwards map which we could use for
  % backwards propagation of approximations.
  An example that is conditionally multiplicative, but does not admit
  a backards map of approximations (from \citet{amadio-curien} just
  before Lemma 12.2.3, originally due to Berry) is given by defining
  $\mathrm{unstable} : D \to \{\bot, \top\}$, where
  $D = \bot \sqsubseteq \cdots \sqsubseteq n \sqsubseteq \cdots
  \sqsubseteq 1 \sqsubseteq 0$, as $\mathrm{unstable}(\bot) = \bot$
  and $\mathrm{unstable}(n) = \top$. This is monotone, and preserves
  meets in every $\downset{x}$. However, there is no ``best'' (i.e.,
  least) input that gives us any finite output.
\end{example}

@@

In light of the \exref{non-stable-function}, we consider
\citet{berry79}'s definition of \emph{stable function} that requires
the existence of a reverse mapping directly, even without assuming
that any meets exist:

@@

  \item As discussed above, $\mathrm{unstable}$ is not stable, but is
    conditionally multiplicative.

  \item The addition function on intervals with nominated points in
    \exref{intervals-and-maxima-elements} is stable. Given the input
    $[l_1,x_1,u_1], [l_2,x_2,u_2]$ and an approximation
    $[l,x_1+x_2,u]$ of the output, the minimal approximations of the
    input are $[l-x_2,x_1,u-x_2], [l-x_1,x_2,u-x_1]$. We can read this
    as saying if the output was the maximal element $x_1+x_2$ but we
    only require the output to be in the range $[l,x_1+x_2,u]$, then
    we can obtain intervals containing the input values that are
    enough to obtain the desired output approximation \emph{assuming
      that the other input is kept the same}. Note the analogy to
    partial derivatives in multi-variable calculus, where the
    derivative is computed in each variable independently.
  \end{enumerate}
\end{example}

% The fact that these two functions' stability witnesses reveal
% information about how they depend on their input is what we will
% exploit in order to use the idea of stability for \GPS. It is
% interesting that sequentiality is not necessarily required in order
% to obtain backwards maps of approximations, as might be assumed by
% thinking about how to trace computations for provenance information.

Stability has an alternative definition in terms of Galois
connections, which will be more useful for what follows. This
characterisation is due to~\citet{taylor99}. We first define Galois
connections, which we used informally in \exref{introduction-example}.

\begin{definition}[Galois connection]
  \label{def:galois-connection}
  Suppose $X$ and $Y$ are posets. A \emph{Galois connection} $f \adj g: X \to Y$ is a pair of monotone functions $f: Y \to X$ and $g: X \to Y$ satisfying $y \leq g(x) \iff f(y) \leq x$ for any $x \in X$ and $y \in Y$. Since a Galois connection is also an adjunction, we refer to $f$ as the left adjoint and $g$ as the right adjoint.
\end{definition}

@@

Even though stable functions can be defined on any partially ordered
set, in light of the analogy with tangent spaces it makes sense to
require that meets, preserved by forward approximation maps, and
joins, preserved by backwards approximation maps, exist:

@@

\begin{theorem}
  L-posets and stable functions form a category with products and
  coproducts.
\end{theorem}

The category of L-posets and stable functions is not cartesian
closed. To make it so, we would need to require that the principal
downsets $\downset{x}$ are \emph{complete}
lattices. \citet{amadio-curien} (Theorem 12.FIXME) details the proof. Intuitively, to generate the best approximation of an input value for a function, we need to take the infimum over all possible input values.

Our goal is to model a higher-order language suitable for writing
queries on databases as in \exref{introduction-example}, so why should
we not just take complete L-posets and stable functions as our model?
We have two reasons for moving to a slightly different model in \secref{model}:

\begin{enumerate}
\item Even without completeness, in Bounded meet posets and L-posets,
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. \bob{fwd reference to discussion.}) In
  \exref{intervals-and-maxima-elements}, maximal elements could be
  taken to be the ``proper values''. One idea is to restrict to
  conditionally multiplicative or stable functions that preserve
  maximal elements. However, this idea fails at higher order:
  functions that take maximal elements to maximal elements are not
  themselves maximal elements. We could devise a category of L-posets
  with totality predicates (which would pick out maximal elements at
  first-order) and totality preserving functions, but we prefer a more
  direct method of separating values proper from their approximations
  using the \emph{Category of Families} construction we detail in the
  next section.
\item A more local reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infinma of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} shows
  how to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

\bob{For future work: what is the connection between FamCPO(LatGal)
  and Laird's bistable biorders, or Berry bidomains?}

% We could have required that L-posets' principal downsets were complete lattices, which would mean that we would only need to require that forward approximation maps preserve all meets and we could compute the backwards map automatically. However, our goal is a computable theory for \GPS, analogous to automatic differentiation. Requiring completeness to compute backwards maps of approximations would require that complete joins are always computable, which will not be the case for approximations of functions.

% \begin{proposition}
%   L-posets and stable functions form a category. \bob{and the main point is that composition of stability witnesses looks like a chain rule!}
% \end{proposition}

% \bob{L-poset is has products and coproducts, but is not closed. To be closed would require each $\downset{x}$ to be a \emph{complete} lattice. However, this is tricky to implement as we need an infinitary operation to compute the reverse approximation map for function types. See \cite{amadio-curien} for the proof that bounded-complete lattices with stable functions form a cartesian closed category.}

@@

We have seen that \citet{berry79}'s theory of stable functions between suitable partial orders can be seen as form of differentiability with forward and reverse derivatives. In the next section, we describe a model based on these ideas suitable for constructing executable models of total higher order languages.

We end this section with a conjecture. Although we have argued above
that there is an analogy between stable functions and smooth
functions, we have not stated any mathematical theorems substantiating
this. \emph{Tangent Categories} \cite{tangent-categories} are a
categorical axiomatisation of the properties of manifolds and smooth
functions in terms of the presence of tangent bundles $TX$ for every
object $X$, forward derivatives and additivity of tangents. They
generalise Cartesian Differential Categories \cite{cdcs}, which are an
axiomatisation of Euclidean spaces and smooth functions.

\begin{conjecture}
  \label{con:tangent-stable-fns}
  (1) Bounded meet posets and conditionally multiplicative functions
  form a Tangent category where the tangent bundle
  $TX = \{(x,x') \mid x' \leq x\}$ and addition of tangents is given
  by meets. (2) L-posets and stable functions form a \emph{reverse
    Tangent category} \cite{reverse-tangents}.
\end{conjecture}

% Let us summarise our analogy between differentiable functions between manifolds and stable functions between L-posets:

% \begin{itemize}
% \item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
% \item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
% \item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
% \item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
% \end{itemize}

% \bob{Put a conjecture about Tangent Categories in somewhere}

% Given this close connection between \GPS and differentiable programming, we can take structures intended for modelling automatic differentiation and use them to model \GPS. This will enable us to generalise and expand the scope of \GPS to act as a foundation for data provenance in a wide range of computational settings.
